Best Practices for Multiple Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-06-13 paper/manuscript.tex

>>>>

\paragraph{Multistate methods}
When dealing with large datasets, methods that compute the binding free energies of multiple ligands simultaneously offer a compelling alternative to pairwise calculations. These approaches -- categorized as "multiple end state" or "multistate" free energy calculations -- sidestep ligand network construction entirely and provide all pairwise free energy differences directly with similar statistical uncertainties (i.e., cycle closure is given by definition). One potential disadvantage of multistate methods is, however, that the presence of multiple end states may complicate the energy landscape, thus requiring longer, inherently sequential simulations. Additionally, they are often more difficult to set up as the combination of multiple ligand topologies into a dual or hybrid topology raises extra challenges \cite{ries2022restraintmaker}.

A well-established method from this category is enveloping distribution sampling (EDS) \cite{christ2007enveloping}, which constructs a single reference state $R$ from the end-states, 
\begin{equation}\label{eq:eds}
H_R (\vec{q}) = - \beta s \ln{ \sum_{i=1}^N \left( e^{-\beta s (H_i(\vec{q})-E_i^R)}\right)},
\end{equation}
where $\vec{q}$ are the coordinates of the system, $\beta=(k_B T)^{-1}$ with $k_B$ being the Boltzmann constant and $T$ the absolute temperature, $N$ is the number of end-states, $s$ is the smoothing parameter and $E^R$ the energy offsets.
The computational efficiency of EDS is mainly due to the fact that interactions between unperturbed protein and solvent degrees of freedom have to be calculated only once independent of $N$, provided a pairwise decomposable method like e.g. reaction field \cite{tironi1995generalized} is used for the long-range electrostatic interactions. Thus, the computational cost of adding more ligands grows approximately linearly rather than quadratically (in a fully connected perturbation graph) \cite{champion2023accelerating}.

To ensure sampling of all end-states in the EDS simulation, the reference-state parameters $s$ and $E^R$ must be optimized.
To simplify the calibration when $N >2$ and to improve the robustness of the methodology, multiple EDS reference-state potentials can be combined in a replica-exchange setup (termed RE-EDS) with different magnitudes of smoothing ($s$-parameter) \cite{sidler2016replica, sidler2017efficient, ries2022relative, champion2023accelerating}. 
RE-EDS simulations can be conducted using either a single or dual coordinate framework \cite{ries2022restraintmaker}. In the first case, all ligands must be restrained together (e.g. with distance restraints) to avoid diffusing out of the binding site. However, the number and definition of restraints should be chosen carefully as they can introduce bias in the simulations if not explicitly corrected. Single coordinate topologies (such as SCDT) can alleviate this issue as well as improve phase-space overlap \cite{champion2023accelerating} although their implementation in most MD engines usually requires custom routines.

RE-EDS implementations in the GROMOS \cite{champion2023accelerating} or OpenMM \cite{rieder2022leveraging} software packages include automated parameter calibration procedures (i.e., optimizing the distribution of replicas in smoothness-parameter space, optimizing the energy offsets). Previous work showed that 15-30 ligands can be treated in a RE-EDS simulations in practice without issues \cite{rieder2022leveraging, champion2023accelerating,champion2024multistate}.
However, sampling issues can occur for ligands that require insertion or deletion of many heavy atoms. In these cases, it is beneficial to split the set of ligands into smaller groups or include such a ligand via pairwise calculations.
